\bf Visual diagnostics for constrained optimisation with application to guided tours

### 2020-10-03 paper/paper.Rmd
change functions in my pacakge to a list

---
title: Title here
blinded: 0
authors:

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

```{r fig.cap = "THis is the caption", fig.height=4, fig.align="center"}
# ani <- p + theme(legend.position = "none") +
#   transition_states(id) +
#   shadow_mark()
# 
# animate(ani, nframes = 350, device = "png",
#         renderer = file_renderer("figures/pca/",
#                                  prefix = "pca", overwrite = TRUE))

@@

Compare to the PCA plot, the animated rotation (tour) display gives a more holistic view of the search of the optimiser and these additional information gained from the animation is cruicial. This is because in essense, PCA presents one reduced space that maximises the variance and it inherently amplifies the spread of the randomly evaluated bases in the search since they have larger variance than the bases on the interpolation path. Hence when the interest is to view the interpolation path in the space, an animated rotation display provides the view of high dimension from different angles.

```{r fig.cap="THis is te caption"}
set.seed(1)
sphere <- geozoo::sphere.hollow(p = 5, n = 1000)$point
path1 <- holes_1d_better$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
path2 <- holes_1d_geo$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
theoretical <- matrix(c(0, 1, 0, 0, 0,
                        0, -1, 0, 0, 0), nrow = 2, byrow = TRUE)
start <- get_start(holes_1d_better) %>% pull(basis) %>% .[[1]] %>% matrix(nrow = 1)
dt <- rbind(sphere, path1, path2, theoretical, start)
colnames(dt) <- c(map_chr(1:5, ~paste0("x", .x)))
pal <- c("#D3D3D3",c("#524340",  #orchre
                     "#B4B754",  # green
                     "#F3B422" # yellow
                     ))
color <- c(rep(pal[1], nrow(sphere)),
           rep(pal[2], nrow(path1)),
           rep(pal[3], nrow(path2)),
           rep(pal[4], 2), 
           pal[2] # for start
           )
cex <- c(rep(1, nrow(sphere)),
         rep(1.5, nrow(path1)), 
         rep(1.5, nrow(path2)), 
         rep(5, 2), 
         5 # for start
         )

@@

2.

@@

>>>>

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
blinded: 0
authors:

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

```{r fig.cap = "\\label{toy-pca-aniamted}A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm.", fig.height=4, fig.align="center"}
# ani <- p + theme(legend.position = "none") +
#   transition_states(id) +
#   shadow_mark()
# 
# ani <- bind_rows(holes_1d_geo, holes_1d_better) %>% 
#   bind_theoretical(matrix(c(0, 1, 0, 0, 0), nrow = 5), 
#                    index = tourr::holes(), raw_data = data) %>% 
#   bind_theoretical(matrix(c(0, -1, 0, 0,0), nrow = 5), 
#                    index = tourr::holes(), raw_data  = data)%>%
#   explore_space_pca(col = method, animate = TRUE)  + 
#   scale_color_botanical(palette = "cherry") 
# 
# animate(ani, nframes = 350, device = "png",
#         renderer = file_renderer("figures/pca/",
#                                  prefix = "pca", overwrite = TRUE))

@@

Compare to the PCA plot, the animated rotation (tour) display, in Figure \ref{toy-tour}, gives a more holistic view of the search of the optimiser and these additional information gained from the animation is cruicial. This is because in essense, PCA presents one reduced space that maximises the variance and it inherently amplifies the spread of the randomly evaluated bases in the search since they have larger variance than the bases on the interpolation path. Hence when the interest is to view the interpolation path in the space, an animated rotation display provides the view of high dimension from different angles.

```{r toy-tour, fig.cap="\\label{toy-tour}A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row view the space from a direction that is close to the one in PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles."}
set.seed(1)
sphere <- geozoo::sphere.hollow(p = 5, n = 1000)$point
path1 <- holes_1d_better$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
path2 <- holes_1d_geo$basis %>% 
  flatten_dbl() %>% matrix(ncol = 5, byrow = TRUE)
theoretical <- matrix(c(0, 1, 0, 0, 0,
                        0, -1, 0, 0, 0), nrow = 2, byrow = TRUE)
start <- get_start(holes_1d_better) %>% pull(basis) %>% .[[1]] %>% matrix(nrow = 1)
dt <- rbind(sphere, path1, path2, theoretical, start)
colnames(dt) <- c(map_chr(1:5, ~paste0("x", .x)))
pal <- c("#D3D3D3",c("#524340",  #orchre
                     "#B4B754",  # green
                     "#F3B422" # yellow
                     ))
color <- c(rep(pal[1], nrow(sphere)),
           rep(pal[2], nrow(path1)),
           rep(pal[3], nrow(path2)),
           rep(pal[4], 2), 
           pal[2] # for start
           )
cex <- c(rep(1, nrow(sphere)),
         rep(1.5, nrow(path1)), 
         rep(1.5, nrow(path2)), 
         rep(5, 2), 
         5 # for start
         )

@@

Once the data object has been obtained, the package, `ferrn`, provides automatic diagnostic plots as described in Section \ref{vis-diag}. The structure of package functionality has been listed below.

**Main functions**

- `explore_trace_search()`: produce summary plots, as shown in Figure \ref{toy-search}
- `explore_trace_interp()`: produce trace plots for the interpolation points, as shown in Figure \ref{toy-interp}
- `explore_space_pca() `: produce plots of projection basis on the reduced space by PCA, as shown in Figure \ref{toy-pca}. Animated version in Figure \ref{toy-pca-aniamted} can be acquired via the argument `animate = TRUE`
- `explore_space_tour()`: produce animated tour view on the full space of the projection bases, as shown in Figure \ref{toy-tour}

**Auxiliary functions**

-  `get_*()` extract and, in some cases, manipulate certain components from the existing data object.
    - `get_best()`: extract the best basis found by the algorithm
    - `get_start()`: extract the starting basis
    - `get_interp()`: extract the observations in the interpolation and prepare to be supplied to `explore_trace_interp()`. 
    - `get_search_count()`: produce the summary table of the number of observation in each iteration and prepare to be supplied to `explore_trace_search()`
    - `get_basis_matrix()`: flattern all the bases into a row and stack them to form a large matrix for `explore_space_tour()`

- `bind_*()` incorporate additional information that is outside the tour optimisation.

  - `bind_theoretical()`: incorporate the best possible basis to the existing data object with the supply of the index function and original data for producing the index value. 
  - `bind_random()`: generates a large number of points on the surface of a high dimensional sphere and attaches it to the existing data object and output as a tibble object. `bind_random_matrix()` returns a matrix.

  <!-- This binding is used in the space plots to compare the position of the theoretical best basis to the ones being searched by the algorithm. The following chunk presents its usage -->
<!-- These are also useful in the space plots since it preserves the spherical structure of the spaces of the projection bases, in the computation of PCA for `explore_space_pca()` and animation of `explore_space_tour()`.  -->

- Color and relevel
  - `botanical_palettes`:  a collection of color palettes from Australian native plants Quantitative palettes include daisy, banksia and cherry and sequantial palettes cotain fern and acacia.
  - `botanical_pal()`: a color interpolator useful for `scale_color_botanical()`
  - `scale_color_botanical()`: a ggplot construction for using botanical palettes. 
  - `relevel_geo()`: manipulate the level in the `info` column in the data object when geodesic search is used
  - `relevel_better()`: manipulate the level in the `info` column in the data object when search_better related search is used

```{r eval = FALSE}
# Below are codes that generate the plots in Section 4 
# using functions from ferrn package
bind_rows(holes_2d_better_max_tries %>% mutate(max_tries = 500), 
          holes_2d_better %>% mutate(max_tries = 25)) %>% 
  explore_trace_search(group = max_tries)
```

```{r eval = FALSE}
bind_rows(holes_2d_better_max_tries %>% mutate(group = "max_tries: 500"), 
          interrupt_no %>% mutate(group = "no interruption"), 
          holes_2d_better_random %>% mutate(group = "better_random")) %>% 
  mutate(group = fct_relevel(group, c("max_tries: 500", 
                                      "no interruption", 
                                      "better_random"))) %>% 
  explore_trace_interp(group = group) + 
  theme(legend.position = "none")
```

```{r eval = FALSE}
bind_rows(holes_1d_geo, holes_1d_better) %>% 
  bind_theoretical(matrix(c(0, 1, 0, 0, 0), nrow = 5), 
                   index = tourr::holes(), raw_data = data) %>% 
  bind_theoretical(matrix(c(0, -1, 0, 0,0), nrow = 5), 
                   index = tourr::holes(), raw_data  = data)%>%
  explore_space_pca(col = method)  + 
  scale_color_botanical(palette = "cherry")
```

```{r eval = FALSE}
bind_rows(holes_1d_geo, holes_1d_better) %>% 
  explore_space_tour(color = method)
```

@@

This paper has illustrated setting up a data object that can be used for diagnosing a complex optimisation procedure. The ideas were illustrated using the optimisers available for projection pursuit guided tour. Here the constraint is the orthornormality condition of the projection bases. The approach used here could be broadly applied to understand other constrained optimisers.

something about the diagnostic plots

summarise what we have done and what might be done in the future.

### 2020-10-03 paper/paper.tex
change functions in my pacakge to a list

\if0\blind
{
  \title{\bf Title here}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title here}
  \end{center}
  \medskip
} \fi

@@

@@

Compare to the PCA plot, the animated rotation (tour) display gives a
more holistic view of the search of the optimiser and these additional
information gained from the animation is cruicial. This is because in
essense, PCA presents one reduced space that maximises the variance and
it inherently amplifies the spread of the randomly evaluated bases in
the search since they have larger variance than the bases on the
interpolation path. Hence when the interest is to view the interpolation
path in the space, an animated rotation display provides the view of
high dimension from different angles.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
\end{enumerate}

>>>>

\if0\blind
{
  \title{\bf Visual diagnostics for constrained optimisation with application to
guided tours}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Visual diagnostics for constrained optimisation with application to
guided tours}
  \end{center}
  \medskip
} \fi

@@

\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-1-1}

}

\caption{\label{toy-pca-aniamted}A selected number of frames from the animated PCA plot. With animation, it is easier to track the progrssion from the start to finish in each algorithm.}\label{fig:unnamed-chunk-1}
\end{figure}

@@

Compare to the PCA plot, the animated rotation (tour) display, in Figure
\ref{toy-tour}, gives a more holistic view of the search of the
optimiser and these additional information gained from the animation is
cruicial. This is because in essense, PCA presents one reduced space
that maximises the variance and it inherently amplifies the spread of
the randomly evaluated bases in the search since they have larger
variance than the bases on the interpolation path. Hence when the
interest is to view the interpolation path in the space, an animated
rotation display provides the view of high dimension from different
angles.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/toy-tour-1.pdf}
\caption{\label{toy-tour}A selected number of frames from the tour
animation for viewing the 5D space of all the projection bases. The
second frame on the top row view the space from a direction that is
close to the one in PCA plot. The tour animation allows for a more
holistic view of the full space in high dimensions from different
angles.}
\end{figure}

@@

Once the data object has been obtained, the package, \texttt{ferrn},
provides automatic diagnostic plots as described in Section
\ref{vis-diag}. The structure of package functionality has been listed
below.

\textbf{Main functions}

\begin{itemize}
\tightlist
\item
  \texttt{explore\_trace\_search()}: produce summary plots, as shown in
  Figure \ref{toy-search}
\item
  \texttt{explore\_trace\_interp()}: produce trace plots for the
  interpolation points, as shown in Figure \ref{toy-interp}
\item
  \texttt{explore\_space\_pca()}: produce plots of projection basis on
  the reduced space by PCA, as shown in Figure \ref{toy-pca}. Animated
  version in Figure \ref{toy-pca-aniamted} can be acquired via the
  argument \texttt{animate\ =\ TRUE}
\item
  \texttt{explore\_space\_tour()}: produce animated tour view on the
  full space of the projection bases, as shown in Figure \ref{toy-tour}
\end{itemize}

\textbf{Auxiliary functions}

\begin{itemize}
\item
  \texttt{get\_*()} extract and, in some cases, manipulate certain
  components from the existing data object.

  \begin{itemize}
  \tightlist
  \item
    \texttt{get\_best()}: extract the best basis found by the algorithm
  \item
    \texttt{get\_start()}: extract the starting basis
  \item
    \texttt{get\_interp()}: extract the observations in the
    interpolation and prepare to be supplied to
    \texttt{explore\_trace\_interp()}.
  \item
    \texttt{get\_search\_count()}: produce the summary table of the
    number of observation in each iteration and prepare to be supplied
    to \texttt{explore\_trace\_search()}
  \item
    \texttt{get\_basis\_matrix()}: flattern all the bases into a row and
    stack them to form a large matrix for
    \texttt{explore\_space\_tour()}
  \end{itemize}
\item
  \texttt{bind\_*()} incorporate additional information that is outside
  the tour optimisation.

  \begin{itemize}
  \tightlist
  \item
    \texttt{bind\_theoretical()}: incorporate the best possible basis to
    the existing data object with the supply of the index function and
    original data for producing the index value.
  \item
    \texttt{bind\_random()}: generates a large number of points on the
    surface of a high dimensional sphere and attaches it to the existing
    data object and output as a tibble object.
    \texttt{bind\_random\_matrix()} returns a matrix.
  \end{itemize}
\item
  Color and relevel

  \begin{itemize}
  \tightlist
  \item
    \texttt{botanical\_palettes}: a collection of color palettes from
    Australian native plants Quantitative palettes include daisy,
    banksia and cherry and sequantial palettes cotain fern and acacia.
  \item
    \texttt{botanical\_pal()}: a color interpolator useful for
    \texttt{scale\_color\_botanical()}
  \item
    \texttt{scale\_color\_botanical()}: a ggplot construction for using
    botanical palettes.
  \item
    \texttt{relevel\_geo()}: manipulate the level in the \texttt{info}
    column in the data object when geodesic search is used
  \item
    \texttt{relevel\_better()}: manipulate the level in the
    \texttt{info} column in the data object when search\_better related
    search is used
  \end{itemize}
\end{itemize}
